---
title: "Workshop 8 TEMPLATE"
subtitle: "9 AM"
author: "Elise LaFreniere"
date: "05-22-2026"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 7 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages-and-data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest")
```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Eschscholzia californica" = esccal_col,
  "Grindelia camporum" = gricam_col,
  "Stipa pulchra" = stipul_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Encelia californica" = enccal_col)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_minimal())
```

# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating new object from drought_exp
drought_exp_clean <- drought_exp |> 
  # cleaning names (column names lowercase with _)
  clean_names() |> 
  # adding full names through creating new column  
  mutate(species_name = recode_values(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # adding full names for water treatments through creating new column 
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  #reordering species levels 
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
 # reordering water treatment levels 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting for desired columns 
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  
 
```{r}
#| label: checking-structure 
# looking to see that fct_revel worked through "Factor w/ x levels output
str(drought_exp_clean)
```

Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: ten-rows 
# selecting rows 
slice_sample(
  drought_exp_clean, # data frame
  n = 10 # selecting for 10 rows 
)
```

# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 
```{r}
#| label: scatterplot 
# base layer : ggplot
ggplot(drought_exp_clean,
       mapping = aes(
         x = sla, # x-axis SLA
         y = total_g # y-axis total mass
       )) +
  # first layer: geom_point, creating scatter plot 
  geom_point() +
  # renaming axes, adding title
  labs(
    x = "Specific Leaf Area (cm\U00B2/g)",
    y = "Total Mass (g)",
    title = "As SLA increases, total mass (g) increases"
  )
```

## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
#| label: jitterplot-water-treatment
# base layer: ggplot
ggplot(drought_exp_clean,
       mapping = aes(
         x = water_treatment, # x-axis water treatment
         y = total_g, # y-axis total mass
        color = water_treatment
         )) + 
  # first layer: geom_jitter
  geom_jitter(height = 0, 
              width = 0.2) +
  # renaming axes, adding title 
  labs(
    x = "Water Treatment",
    y = "Total Mass (g)",
    title = "On Average, well watered plants have higher mass\nthan drought stressed"
  ) +
  # removing theme
  theme(legend.position = "none",
        plot.title.position = "plot") 
```

## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.

```{r}
#| label: jitter-species
# base plot: ggplot
ggplot(drought_exp_clean,
       mapping = aes(
         x = species_name,
         y = total_g,
         color = species_name 
       )) + 
  # first layer: geom jitter to represent plant individuals 
  geom_jitter(height = 0, 
              width = 0.2) +
  # naming axes, adding title 
  labs(
    x = "Species Name",
    y = "Total Mass (g)",
    title = "Species differ in total plant mass"
  ) +
  # wrapping x axis text
  scale_x_discrete(labels = label_wrap(10)) +
  # remove legend 
  theme(legend.position = "none") + 
  # using custom colors assigned in object
  scale_color_manual(values = plant_colors
  )
```

# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model

model0 <- lm(
  total_g ~ 1, # formula no predictors
  data = drought_exp_clean # clean data frame
)

# model 1: all predictors

model1 <- lm(
  total_g ~ sla + water_treatment + species_name, # formula all predictors
  data = drought_exp_clean # clean data frame
)

# model 2: SLA and water treatment

model2 <- lm(
  total_g ~ sla + water_treatment, # formula sla and water treatment 
  data = drought_exp_clean  # clean data frame
)

# model 3: SLA and species

model3 <- lm(
  total_g ~ sla + species_name, # formula sla and species name 
  data = drought_exp_clean  # clean data frame
)
# model 4: water treatment and species

model4 <- lm(
  total_g ~ water_treatment + species_name,# formula water treatment and species name 
  data = drought_exp_clean  # clean data frame
)

# model 5: SLA

model5 <- lm(
  total_g ~ sla, # formula sla 
  data = drought_exp_clean  # clean data frame
)

# model 6: water treatment

model6 <- lm(
  total_g ~ water_treatment, # formula water treatment 
  data = drought_exp_clean  # clean data frame
)

# model 7: species

model7 <- lm(
  total_g ~ species_name, # formula species name 
  data = drought_exp_clean  # clean data frame
)
```

## Model diagnostics


```{r}
# displaying plots in 2 x 2 grid 
par(mfrow = c(2,2))
# generating diagnostic plots
plot(model4)
```

- residuals appear normally distributed (qq residuals)
- residuals appear homoscedastic (constant variance)
- no clear outliers in data set influencing model coefficient estimates

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: calc-AIC
# calculating AIC from models 
AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7) |> 
  # rearranged to lowest to highest values 
  arrange(AICc)
```



Insert code to display the summary for the best model. 

```{r}
#| label: model-summary 
# summary function for chosen model
summary(model4)
```

- `(Intercept)`: model prediction for plant mass for *Lotus scorparius* in the drought stressed treatment (because those are the reference levels)
- `water_treatmentWell watered` : model prediction for difference in *Lotus scorparius* mass between the well watered and drought stressed treatment
- `species_namePenstemon centranthifolius` : model prediction for the difference between Penstemon mass in the well watered treatment and Lotus mass in well watered treatment 
- `species_nameEschscholzia californica` :  model prediction for the difference between Eschscholzia mass in the well watered treatment and Lotus mass in well watered treatment 



## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors.

**How much variation in the response (total mass, in grams) does this model explain?**

72.6% of the variation in plant mass is explained by a model with both of these predictors.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

- Treatment influences plant mass. 
- Species influences plant mass.

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: model-predictions 

preds <- ggpredict( 
  model4,
  terms = c("water_treatment", "species_name")) |> 
  rename(water_treatment = x, 
         species_name = group)
```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
ggplot(data = drought_exp_clean,
       mapping = aes(x = water_treatment, 
                     y = total_g,
                     color = water_treatment)) + 
  geom_jitter(height = 0, 
              width = 0.1,
              shape = 21, 
              alpha = 0.8) +
  geom_pointrange(data = preds,
                  mapping = aes(x = water_treatment,
                                y = predicted,
                                ymin = conf.low,
                                ymax = conf.high)) +
  facet_wrap(~ species_name) + 
  theme(legend.position = "none")




```


# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







